Monte Carlo as Approximate Polynomial Preconditioning within MCSA

### 2014-05-19 doc/polynomial_note/polynomial.tex
Adding results on shifted Laplacian.

As a simple test problem, we choose the ``JPWH 991'' matrix from
the Harwell-Boeing Collection, available from the Matrix Market.
The sparsity pattern of the JPWH matrix is given in Figure~\ref{fig:jpwh_sparsity}.
\begin{figure}
\centering
\includegraphics[width=4.5in]{jpwh_sparsity}
\caption{Sparsity pattern of JPWH matrix \label{fig:jpwh_sparsity}}
\end{figure}
The matrix is nonsymmetric, but its eigenvalues are all real and
contained in the interval $[-16.3,-0.12]$.  After a diagonal row
scaling, the eigenvalues are contained in the interval $[0.02,1.71]$.
All test cases use the diagonally scaled matrix and the right
hand side of the linear system is taken to be a vector of all ones.
The stopping criterion is a reduction in the L$_2$ norm of the residual
by six orders of magnitude.

Table~\ref{tab:jpwh_standard_neumann} shows the behavior of
the standard MCSA algorithm applied to the JPWH matrix for a range
of different polynomial orders and Monte Carlo histories per iteration.
As expected, increasing the polynomial order generally leads to a decrease
in the number of iterations required for convergence.
For a fixed number of histories, however, there is a threshold beyond
which an increase in the polynomial order actually leads to very slow
convergence or even divergence.  Conversely, for a fixed polynomial order
it is necessary to track ``enough'' histories to ensure stability, but
beyond that point there is little benefit to executing additional histories.
\begin{table}
\caption{Standard MCSA with Neumann Basis, JPWH Matrix.
\label{tab:jpwh_standard_neumann}}
\centering
\begin{tabular}{cccccccccc}
\toprule
Histories per & \multicolumn{9}{c}{Polynomial Order} \\
Iteration & 1 & 2 & 4 & 8 & 16 & 32 & 64 & 128 & 256 \\
\midrule
100 & 234 & 175 & 121 & 183 & - & - & - & - & - \\
200 & 230 & 173 & 118 & 74 & 164 & - & - & - & - \\
400 & 227 & 171 & 115 & 71 & 42 & 91 & - & - & - \\
800 & 227 & 170 & 114 & 70 & 39 & 31 & 47 & 67 & 78 \\
1600 & 226 & 170 & 113 & 69 & 39 & 22 & 22 & 27 & 27 \\
3200 & 225 & 169 & 113 & 68 & 38 & 21 & 15 & 17 & 17 \\
6400 & 225 & 169 & 113 & 68 & 38 & 21 & 12 & 12 & 13 \\
12800 & 225 & 169 & 113 & 68 & 38 & 21 & 11 & 10 & 10 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:jpwh_cheby_power} shows the same result,
but now instead of using $\bI - \bA$ as the basis for the
polynomial expansion, $\bA$ is used instead.
Thus, instead of Eq.~\eqref{eq:mc_decomp} defining the
matrices to be used in the Monte Carlo process, we instead
have
\begin{equation}
\bA^T = \bP \circ \bW \:.
\end{equation}
We see that
for this basis, the majority of cases result in divergence.
Interestingly, at low polynomial orders and large history
counts, the behavior is nearly identical to that with
$\bI - \bA$ as the basis.
\begin{table}
\caption{Standard MCSA with Monomial Basis, JPWH Matrix.
\label{tab:jpwh_standard_power}}
\centering
\begin{tabular}{cccccccccc}
\toprule
Histories per & \multicolumn{9}{c}{Polynomial Order} \\
Iteration & 1 & 2 & 4 & 8 & 16 & 32 & 64 & 128 & 256 \\
\midrule
    100 & -     & -     & - & - & - & - & - & - & - \\
    200 & 330 & -     & - & - & - & - & - & - & - \\
    400 & 240 & -     & - & - & - & - & - & - & - \\
    800 & 235 & 290 & - & - & - & - & - & - & - \\
  1600 & 230 & 182 & - & - & - & - & - & - & - \\
  3200 & 228 & 177 & - & - & - & - & - & - & - \\
  6400 & 227 & 174 & - & - & - & - & - & - & - \\
12800 & 226 & 172 & - & - & - & - & - & - & - \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:jpwh_cheby_neumann} shows the performance
with Chebyshev polynomials rather than the Neumann
series.  Note that the polynomial order is increased
linearly in this case rather than exponentially with the
Neumann series.  In general, using Chebyshev polynomials
results in a significant reduction in iterations relative to
the equivalent order Neumann series polynomial.  The
downside is that significantly more histories must be
tracked to achieve stability for a given polynomial order.
\begin{table}
\caption{Chebyshev MCSA with Neumann Basis, JPWH Matrix.
\label{tab:jpwh_cheby_neumann}}
\centering
\begin{tabular}{cccccc}
 \toprule
 Histories per & \multicolumn{5}{c}{Polynomial Order} \\
 Iteration & 1 & 2 & 3 & 4 & 5 \\
 \midrule
 100 & - & - & - & - & - \\
 200 & 156 & - & - & - & - \\
 400 & 131 & - & - & - & - \\
 800 & 126 & 96 & - & - & - \\
 1600 & 125 & 67 & - & - & - \\
 3200 & 124 & 65 & 85 & - & - \\
 6400 & 123 & 64 & 42 & - & - \\
 12800 & 123 & 63 & 40 & 87 & - \\
 25600 & 123 & 63 & 40 & 33 & - \\
 51200 & 123 & 63 & 39 & 28 & 100 \\
 102400 & 122 & 63 & 39 & 28 & 33 \\
 204800 & 122 & 63 & 39 & 27 & 22 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:jpwh_cheby_power} shows the result of
again using Chebyshev polynomials, but now using
$\bA$ as the basis for the polynomial expansion.
As with the Neumann series polynomial, using
$\bA$ as a basis performs considerably worse than
using $\bI - \bA$ in that considerably more histories
are required to achieve stability for a given polynomial
order.
\begin{table}
\caption{Chebyshev MCSA with Monomial Basis, JPWH Matrix.
\label{tab:jpwh_cheby_power}}
\centering
\begin{tabular}{cccccc}
 \toprule
 Histories per & \multicolumn{5}{c}{Polynomial Order} \\
 Iteration & 1 & 2 & 3 & 4 & 5 \\
 \midrule
 100 & - & - & - & - & - \\
 200 & - & - & - & - & - \\
 400 & - & - & - & - & - \\
 800 & - & - & - & - & - \\
 1600 & 178 & - & - & - & - \\
 3200 & 130 & - & - & - & - \\
 6400 & 127 & - & - & - & - \\
 12800 & 125 & - & - & - & - \\
 25600 & 124 & 80 & - & - & - \\
 51200 & 123 & 67 & - & - & - \\
 102400 & 123 & 65 & - & - & - \\
 204800 & 123 & 64 & - & - & - \\
\bottomrule
\end{tabular}
\end{table}

>>>>

All test cases use the diagonally scaled matrix and the right
hand side of the linear system is taken to be a vector of all ones.
The stopping criterion is a reduction in the L$_2$ norm of the residual
by six orders of magnitude.

Table~\ref{tab:jpwh_standard_neumann} shows the behavior of
the standard MCSA algorithm applied to the JPWH matrix for a range
of different polynomial orders and Monte Carlo histories per iteration.
As expected, increasing the polynomial order generally leads to a decrease
in the number of iterations required for convergence.
For a fixed number of histories, however, there is a threshold beyond
which an increase in the polynomial order actually leads to very slow
convergence or even divergence.  Conversely, for a fixed polynomial order
it is necessary to track ``enough'' histories to ensure stability, but
beyond that point there is little benefit to executing additional histories.
\begin{table}
\caption{Forward MCSA with Neumann Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
Values are MCSA iteration counts (timing in milliseconds)
\label{tab:lap_forward_neumann}}
\centering
\begin{tabular}{ccccccc}
\toprule
Histories per & \multicolumn{6}{c}{Polynomial Order} \\
Iteration & 1 & 2 & 3 & 4 & 5 & 6 \\
\midrule
250 & - & - & - & - & - & - \\
500 & - & - & - & - & - & - \\
1000 & - & - & - & - & - & - \\
2000 & - & - & - & - & - & - \\
4000 & - & 765(249) & 605(285) & 947(560) & 396(302) & 560(458) \\
8000 & 923(325) & 698(413) & 554(485) & 471(560) & 396(530) & 354(561) \\
16000 & 923(616) & 695(809) & 554(946) & 463(1015) & 396(1068) & 348(1079) \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\caption{Adjoint MCSA with Neumann Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
Values are MCSA iteration counts (timing in milliseconds)
\label{tab:lap_adjoint_neumann}}
\centering
\begin{tabular}{ccccccc}
\toprule
Histories per & \multicolumn{6}{c}{Polynomial Order} \\
Iteration & 1 & 2 & 3 & 4 & 5 & 6 \\
\midrule
250 & - & 775(100) & - & 651(96) & - & 664(110) \\
500 & - & 725(122) & - & 502(104) & - & 394(95) \\
1000 & - & 707(174) & - & 482(179) & - & 366(144) \\
2000 & 2096(521) & 703(280) & 1991(855) & 471(259) & 2668(1514) & 356(251) \\
4000 & 1694(821) & 698(494) & 1701(1194) & 467(497) & 864(921) & 350(458) \\
8000 & 1384(1142) & 697(923) & 1256(1761) & 464(905) & 628(1290) & 350(873) \\
16000 & 1220(1969) & 695(1796) & 859(2493) & 462(1768) & 588(2384) & 347(1711) \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\caption{Forward MCSA with Chebyshev Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
Values are MCSA iteration counts (timing in milliseconds)
\label{tab:lap_forward_cheby}}
\centering
\begin{tabular}{ccccccc}
\toprule
Histories per & \multicolumn{6}{c}{Polynomial Order} \\
Iteration & 1 & 2 & 3 & 4 & 5 & 6 \\
\midrule
250 & - & - & - & - & - & - \\
500 & - & - & - & - & - & - \\
1000 & - & - & - & - & - & - \\
2000 & - & - & - & - & - & - \\
4000 & - & - & - & - & - & - \\
8000 & - & - & - & - & - & - \\
16000 & 556(369) & - & - & - & - & - \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\caption{Adjoint MCSA with Chebyshev Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
Values are MCSA iteration counts (timing in milliseconds)
\label{tab:lap_adjoint_cheby}}
\centering
\begin{tabular}{ccccccc}
\toprule
Histories per & \multicolumn{6}{c}{Polynomial Order} \\
Iteration & 1 & 2 & 3 & 4 & 5 & 6 \\
\midrule
250 & - & - & - & - & - & - \\
500 & - & - & - & - & - & - \\
1000 & - & - & - & - & - & - \\
2000 & - & 325(130) & - & - & - & - \\
4000 & - & 294(210) & - & - & - & - \\
8000 & - & 288(380) & 201(335) & - & - & - \\
16000 & - & 285(736) & 175(554) & - & - & - \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\caption{Forward MCSA with GMRES Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
Values are MCSA iteration counts (timing in milliseconds)
\label{tab:lap_forward_gmres}}
\centering
\begin{tabular}{ccccccc}
\toprule
Histories per & \multicolumn{6}{c}{Polynomial Order} \\
Iteration & 1 & 2 & 3 & 4 & 5 & 6 \\
\midrule
250 & - & - & - & - & - & - \\
500 & - & - & - & - & - & - \\
1000 & 923(89) & - & - & - & - & - \\
2000 & 923(127) & 547(110) & - & - & - & - \\
4000 & 923(195) & 547(204) & - & - & - & - \\
8000 & 923(330) & 547(322) & 357(312) & - & - & - \\
16000 & 923(607) & 547(640) & 357(607) & 246(536) & - & - \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\caption{Adjoint MCSA with GMRES Polynomial, $1000 \times 1000$ Shifted Laplacian Matrix.
Values are MCSA iteration counts (timing in milliseconds)
\label{tab:lap_adjoint_gmres}}
\centering
\begin{tabular}{ccccccc}
\toprule
Histories per & \multicolumn{6}{c}{Polynomial Order} \\
Iteration & 1 & 2 & 3 & 4 & 5 & 6 \\
\midrule
250 & 1153(136) & 2572(337) & - & - & - & - \\
500 & 1003(151) & 616(105) & 462(87) & - & - & - \\
1000 & 946(226) & 566(142) & 385(110) & 496(161) & - & - \\
2000 & 933(306) & 557(222) & 368(175) & 260(146) & - & - \\
4000 & 928(519) & 554(412) & 362(311) & 254(259) & 203(236) & - \\
8000 & 924(971) & 549(731) & 362(585) & 251(490) & 184(423) & - \\
16000 & 923(1816) & 548(1426) & 360(1135) & 247(965) & 179(797) & 146(744) \\
\bottomrule
\end{tabular}
\end{table}
